Precise spiking motifs in neurobiological and neuromorphic data

### 2022-11-08 manuscript.tex
1

However, the performance of SNNs is still lagging compared to that of analog networks and the question of the advantage of using spikes in machine learning and computer vision remains open. Achieving to efficiently implement this type of modeling is extremely important with respect to the development of a new generation of cameras called Silicon Retinas. These represent visual information as events, similarly to the biologically-inspired representation described above and in contrast with a basic frame-based representation~\citep{lichtsteiner_128-128_2008} (for a review, see~\citep{rasetto_challenges_2022}, this will be developped in Section~\ref{sec:neuromorphic}). % This work has been extended to include unsupervised learning capabilities, and we have recently developed a SNN architecture that allows to categorize images of different classes in only a few spikes~\citep{grimaldi_homeostatic_2021,grimaldi_robust_2022}. 
To alleviate this gap in computational neuroscience, we will further review the role of  precise spike timing in neural assemblies from a neuroscientific perspective in the next section.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Role of precise spike timing in neural assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{One first hypothesis: synchronous firing in cell assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
In his book ``Corticonics'',~\citet{abeles_corticonics_1991} queried whether the role of cortical neurons is to integrate synaptic inputs or rather to detect coincidences in temporal spiking patterns. The book gradually leads the reader from the macroscopic cortical anatomy and standard electro-physiological properties of single neurons to neural network models. While the first hypothesis favors the rate coding theory, the second possibility highlights the need for temporal precision in the neural code that we have already explored at the cellular level~\citep{abeles_role_1982,paugam-moisy_computing_2012}. The book then demonstrates that this could take the form of ``synfire chains,'' that is, the emergence of synchronous activity on subsets of neurons which could be propagated in a stable fashion. Since this date, multiple experimental observations have suggested the existence of this precise zero-phase-lag spike synchronization in a defined subset of neurons~\citep{harris_organization_2003}. One possible function of this synchronization may serve the binding of information distributed in the brain~\citep{singer_visual_1995, roelfsema_visuomotor_1997}.

Some experimental results show the emergence of synchrony, for instance in motor cortical function~\citep{riehle_spike_1997}. Interestingly, the authors showed that ``Accurate spike synchronization occurred in relation to external events (stimuli, movements) and was commonly accompanied by discharge rate modulations but without precise time locking of the spikes to these external events. Spike synchronization also occurred in relation to purely internal events (stimulus expectancy), where firing rate modulations were distinctly absent. These findings indicate that internally generated synchronization of individual spike discharges may subserve the cortical organization of cognitive motor processes.'' Moreover, such emergence could change over the learning period involved in learning a task~\citep{kilavik_long-term_2009} and showed some tuning to movement direction and reaction time~\citep{grammont_spike_2003}. It is important to note that synchronous events tend to lock to spatio-temporal patterns of neural activity called \href{https://en.wikipedia.org/wiki/Local_field_potential}{LFP} beta waves~\citep{denker_lfp_2018}, and were extended to larger assemblies~\citep{torre_synchronous_2016, torre_asset_2016} using statistical methods (see Section~\ref{sec:detection} for further details). Note also that synchronicity may explain some unintuitive results. Indeed, it has been shown that thalamo-cortical synapses are relatively weak compared to the amount of intra-cortical activity. However, this pathway is sufficient to drive the cortex, as this input is more often synchronously active~\citep{bruno_cortex_2006}.

@@

More generally, the idea of using the synchrony of co-activation in a cell assembly is reminiscent of the hypothesis that was formalized by~\citet{hebb_organization_1949}. 
This is also expressed in the idea that different cortical areas could achieve binding by synchrony~\citep{fries_mechanism_2005}. The synchronicity will generate rhythms at different range of frequencies, with spikes arriving at peak susceptibility (top of a cycle) or down. Such a theory has surprisingly been validated in EEG recordings to explain for example the continuous wagon-wheel illusion, i.e. the perceived reversal of the rotational movements of the spikes in a rotating wheel~\citep{vanrullen_continuous_2006}. More generally, it can be shown that the phase of alpha oscillations (about $10$ Hz) is causally linked with modulations of cortical excitability and with visual perception~\citep{dugue_phase_2011}. 
%
\subsection{A further hypothesis: traveling waves}
%
To further investigate the role of precise timing, let us also focus on the role of differential timings in an assembly of neurons. As we have seen, a visual feature will induce the firing of different cells at different latencies~\citep{celebrini_dynamics_1993}. Using intracellular recordings, it was shown that the response to a focal visual activation would elicit a latency basin, that is, a graded onset of the neural response from the most activated to neighboring neurons~\citep{bringuier_horizontal_1999}. In particular, it was shown that the network of the so-called horizontal connections within a cortical area is typically not myelinated and that this latency basin would be determined by the propagation speed within that area. For generic visual scenes, these processes would generate a complex interplay between the dynamics of the sensory signal and the spatio-temporal determinants of these interactions and may underlay anticipatory mechanisms in the primary visual cortex~\citep{benvenuti_anticipatory_2020,le_bec_horizontal_2022}.
%

>>>>

However, the performance of SNNs is still lagging compared to that of analog networks and the question of the advantage of using spikes in machine learning and computer vision remains open. Achieving to efficiently implement this type of modeling is extremely important with respect to the development of a new generation of cameras called Silicon Retinas. These represent visual information as events, similarly to the biologically-inspired representation described above and in contrast with a basic frame-based representation~\citep{lichtsteiner_128x128_2008} (for a review, see~\citep{rasetto_challenges_2022}, this will be developped in Section~\ref{sec:neuromorphic}). % This work has been extended to include unsupervised learning capabilities, and we have recently developed a SNN architecture that allows to categorize images of different classes in only a few spikes~\citep{grimaldi_homeostatic_2021,grimaldi_robust_2022}. 
To alleviate this gap in computational neuroscience, we will further review the role of  precise spike timing in neural assemblies from a neuroscientific perspective in the next section.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Role of precise spike timing in neural assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{One first hypothesis: synchronous firing in cell assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
In his book ``Corticonics'',~\citet{abeles_corticonics_1991} queried whether the role of cortical neurons is to integrate synaptic inputs or rather to detect coincidences in temporal spiking patterns. The book gradually leads the reader from the macroscopic cortical anatomy and standard electro-physiological properties of single neurons to neural network models. While the first hypothesis favors the rate coding theory, the second possibility highlights the need for temporal precision in the neural code~\citep{abeles_role_1982,paugam-moisy_computing_2012}. The book then demonstrates that this could take the form of ``synfire chains,'' that is, the emergence of synchronous activity on subsets of neurons which could be propagated in a stable fashion. More broadly, the idea of using the synchrony of co-activation in a cell assembly is reminiscent of the hypothesis that was formalized by~\citet{hebb_organization_1949}. Since this date, multiple experimental observations have suggested the existence of this precise zero-phase-lag spike synchronization in a defined subset of neurons~\citep{harris_organization_2003}. One possible function of this synchronization may serve the binding of information distributed in the brain~\citep{singer_visual_1995, roelfsema_visuomotor_1997}.

Some experimental results show the emergence of synchrony, for instance in motor cortical function~\citep{riehle_spike_1997}. Interestingly, these authors showed that ``accurate spike synchronization occurred in relation to external events (stimuli, movements) and was commonly accompanied by discharge rate modulations but without precise time locking of the spikes to these external events. Spike synchronization also occurred in relation to purely internal events (stimulus expectancy), where firing rate modulations were distinctly absent. These findings indicate that internally generated synchronization of individual spike discharges may subserve the cortical organization of cognitive motor processes.'' Moreover, such emergence could change over the learning period involved in learning a task~\citep{kilavik_long-term_2009} and showed some tuning to movement direction and reaction time~\citep{grammont_spike_2003}. It is important to note that synchronous events tend to lock to spatio-temporal patterns of neural activity called \href{https://en.wikipedia.org/wiki/Local_field_potential}{LFP} beta waves~\citep{denker_lfp_2018}, and were also extended to larger assemblies~\citep{torre_synchronous_2016, torre_asset_2016} using statistical methods (see Section~\ref{sec:detection} for further details). Note also that synchronicity may explain some unintuitive results. Indeed, it has been shown that thalamo-cortical synapses are relatively weak compared to the amount of intra-cortical activity. However, this pathway is sufficient to drive the cortex, as this input is more often synchronously active~\citep{bruno_cortex_2006}.

@@

This is also expressed in the idea that different cortical areas could achieve binding by synchrony~\citep{fries_mechanism_2005}. The synchronicity will generate rhythms at different range of frequencies, with spikes arriving at peak susceptibility (top of a cycle) or down. Such a theory has surprisingly been validated in EEG recordings to explain for example the continuous wagon-wheel illusion, i.e. the perceived reversal of the rotational movements of the spikes in a rotating wheel~\citep{vanrullen_continuous_2006}. More generally, it can be shown that the phase of alpha oscillations (about $10$ Hz) is causally linked with modulations of cortical excitability and with visual perception~\citep{dugue_phase_2011}. 
%
\subsection{A further hypothesis: traveling waves}
%
To further investigate the role of precise timing, let us also focus on the role of differential timings in an assembly of neurons. As we have seen, a visual feature will induce the firing of different cells at different latencies~\citep{celebrini_dynamics_1993}. Using intracellular recordings, it was shown that the response to a focal visual activation would elicit a latency basin, that is, a graded onset of the neural response from the most activated to neighboring neurons~\citep{bringuier_horizontal_1999}. In particular, it was shown that the network of the so-called horizontal connections within a cortical area is typically not myelinated and that this latency basin would be determined by the propagation speed within that area. For generic visual scenes, these processes would generate a complex interplay between the dynamics of the sensory signal and the spatio-temporal determinants of these interactions and may underlay anticipatory mechanisms in the primary visual cortex~\citep{benvenuti_anticipatory_2020,le_bec_horizontal_2022}.
%
